---
title: "Retrospective strength of schedule in the NBA (shiny app)"
output: github_document
---

```{r load, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# Clear Workspace
rm(list = ls())
options(scipen = 999)
options(stringsAsFactors = FALSE)

# load packages
library(formatR)
library(tidyverse)
library(ggthemes)
library(lubridate)
library(ggrepel)
local <- TRUE

# read data
if (local) {
    eloDat <- read.table("538elo.csv", header = TRUE, dec = ".", sep = ";")
    rm(local)
} else {
    url538 <- "https://projects.fivethirtyeight.com/nba-model/nba_elo.csv"
    eloDat <- read.table(url538, header = TRUE, sep = ",", dec = ".", na.strings = "")
    eloDat$date <- ymd(eloDat$date)
}

# Make conference dat
confDat <- data.frame(team1 = sort(unique(eloDat$team1[eloDat$season >= 1977])), conference = c(rep("East", 
    8), "West", "West", "East", "West", "West", "East", "West", rep("West", 3), "East", "East", "West", 
    "East", "East", "West", "West", "East", "East", "West", "East", "East", rep("West", 6), "East", "West", 
    "West", "East", "East"))

# Pre-Calc data
tempElo <- eloDat %>% mutate(playoffgame = !is.na(playoff)) %>% group_by(season, team1) %>% summarise_if(is.logical, 
    sum) %>% right_join(eloDat, by = c("season", "team1")) %>% mutate(playoffteam = playoffgame > 0) %>% 
    filter(season >= 1977) %>% left_join(confDat, by = "team1") %>% mutate(playoffgame = !is.na(playoff))

```


During the regular season a common measure of a teams remaining strength of schedule (SOS) is the combined winning percentage of 
those opponents. [link](http://www.tankathon.com/remaining_schedule_strength). A high winning percentage means the opponents you still 
have to face are probably tough and vice-versa. 

There are two ideas I want to apply in this work that are in a similar vein:

1. Substitute the winning percentage with a more adequate measure of performance: the Elo score.
2. Instead of a *remaining* SOS, calculate a *retrospective* SOS.

The first addition is self explanatory and through 538s work very easy to implement. The latter idea 

## Data

* Data since season 1977 (NBA/ABA Merger)

## Results

The variation in Opponents Elo is a lot smaller than the general Team Elo (i.e. the strength of teams and not their opponents). 
That is to be expected as the strength of an opponent can only vary so much through a season.

```{r sd, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
tempElo %>% filter(is.na(playoff)) %>% group_by(season, team1) %>% summarise_at(vars(elo1_pre, 
    elo2_pre), mean) %>% ungroup() %>% group_by(season) %>% summarise_at(vars(elo1_pre, elo2_pre), sd, 
    na.rm = TRUE) %>% ungroup() %>% ggplot(aes(x = season, y = elo2_pre)) + geom_point() + geom_line() + 
    theme_minimal() + geom_rangeframe(sides = "l")+xlab("Season")+ylab("Variation in season-wise opponents Elo")
tempElo %>% filter(is.na(playoff)) %>% group_by(season, team1) %>% summarise_at(vars(elo1_pre, 
    elo2_pre), mean) %>% ungroup() %>% group_by(season) %>% summarise_at(vars(elo1_pre, elo2_pre), sd, 
    na.rm = TRUE) %>% ungroup() %>% ggplot(aes(x = season, y = elo1_pre)) + geom_point() + geom_line() + 
    theme_minimal() + geom_rangeframe(sides = "l")+xlab("Season")+ylab("Variation in season-wise Team Elo")
```


## To-Do

* [x] Upload shiny to shinyapps.io
* [x] Add more seasons (by solving conferences )
* [x] Add carmElo for the recent seasons (via button)
* [ ] Make the graphic bigger in browser
* [ ] Look into the max/min delta a team can undergo in a season
* [x] Visualize the variation in oppoElo and teamElo
* [x] Make multiple seasons selectable
* [ ] Replace team names if it's only name change
* [ ] How similar are the different Elo scores (logreg and visual)